Stability in human interaction networks: primitive typology of vertex, prominence of measures and activity statistics

### 2015-02-15 paper.tex
padronizados nomes dos autores
adicionadas versoes anotadas das revisoes do chu
feita revisao ateh boa parte dos metodos

\author{Renato Fabbri}%
 \homepage{http://ifsc.usp.br/~fabbri/}
 \email{fabbri@usp.br}
  \affiliation{ 
S\~ao Carlos Institute of Physics, University of S\~ao Paulo (IFSC/USP)%\\This line break forced with \textbackslash\textbackslash
}%
%
%\author{Camilo Akimushkin Valencia}
%  \homepage{http://ifsc.usp.br/}
%  \email{akimushkin@ifsc.usp.br}
%  \altaffiliation[Also at ]{IFSC-USP}%Lines break automatically or can be forced with \\
%

\author{Vilson Vieira da Silva Junior}
  \homepage{http://automata.cc/}
  \email{vilson@void.cc}
  \altaffiliation[Also at ]{IFSC-USP}%Lines break automatically or can be forced with \\

@@

\author{Deborah Christina Antunes}
  \homepage{http://lattes.cnpq.br/1065956470701739}
  \email{deborahantunes@gmail.com}
  \altaffiliation{
Curso de Psicologia, Universidade Federal do Cer\'a (UFC)
}%Lines break automatically or can be forced with \\

\author{Marilia Mello Pisani}
  \homepage{http://lattes.cnpq.br/6738980149860322}
  \email{marilia.m.pisani@gmail.com}
 \altaffiliation{
Centro de Ciências Naturais e Humanas, Universidade Federal do ABC (CCNH/UFABC)
}%Lines break automatically or can be forced with \\

%\author{Luciano da Fontoura Costa}
%  \homepage{http://cyvision.ifsc.usp.br/~luciano/}
%  \email{ldfcosta@gmail.com}
%  \altaffiliation[Also at ]{IFSC-USP}%Lines break automatically or can be forced with \\
%
\author{Osvaldo N. Oliveira Jr.}
  \homepage{www.polimeros.ifsc.usp.br/professors/professor.php?id=4}
  \email{chu@ifsc.usp.br}
 \altaffiliation[Also at ]{IFSC-USP}%Lines break automatically or can be forced with \\

@@

\begin{abstract}
    This article reports a characterization of interaction networks and its temporal stability. Such a task involves a selection of aspects to investigate, which lead to: 1) activity distribution in time and among participants; 2) a sound classification of vertex: peripheral, intermediary and hub sectors; 3) combination of basic measures into components with greater dispersion (PCA). While time patterns of activity are not obvious, participant activity follows concentrations expected in scale-free networks. Comparison of incident networks with ideal Erd\"os-R\'enyi networks bearing the same number of edges and vertexes reveals a sound criterion for distinguishing sectors on the networks. Principal components in the basic measures space revealed interesting and regular patterns of independence and dispersion. This includes a ranking of measures that most contribute to dispersion: 1) degree and strength measures, 2) symmetry related quantization, and 3) clusterization. Results suggest typologies for these networks and participants. Further work include considerations of text production, psychoanalysis inspired typologies, participatory democracy exploitation of observed properties, and better visualization support for network evolution.
\end{abstract}

\pacs{89.75.Fb,05.65.+b,89.65.-s}% PACS, the Physics and Astronomy
\keywords{complex networks, social network analysis, pattern recognition, statistics}%Use showkeys class option if keyword
\maketitle

\begin{quotation}
% `The reason for the persistent plausibility of the typological approach, however, is not a static biological one, but just the opposite: dynamic and social. The fact that human society has been up to now divided into classes affects more than the external relations of men. The marks of social repression are left within the individual soul.' 
% 
`The conception of personality structure is the
best safeguard against the
inclination to attribute persistent trends in the
individual to something
"innate" or "basic" or "racial" within him. The
Nazi allegation that natural, biological traits decide the total being of a person
would not have been such
a successful political device
had it not been possible to point to numerous
instances of relative fixity in human behavior and to
challenge those who
thought to explain them on any basis other than a biological one.'

\emph{- Adorno et al, 1969, p. 747}
\end{quotation}

@@

\subsection{Related work}
This work approaches network stability through temporal evolution observations.
This evolution is characterized by a constant number of contiguous messages, of which snapshots are considered to yield a timeline of the network, which is not explored to date, as far as authors know~\cite{netWindowEvo}.
 Works on network evolution often consider solely network growth, in which there is a monotonic increase in the number of events considered~\cite{barabasiEvo}. Moreover, selected exceptions are reported in this section.

 The evolution of interaction networks was addressed with a community focus, in a work that ignores the direction of edges~\cite{barabasiEvo}. Two topologically different networks are reported to emerge, depending on the frequency of interactions, presenting a generalized power law or an exponential connectivity distribution~\cite{barabasiTopologicalEv}. 
In email networks, free-scale properties were verified~\cite{bird}, and different linguistic traces were related to weak and strong ties~\cite{GMANE2}.

 Such results are in accordance with phenomena observed in this work. Linguistic characterization is being described in a deriving article~\cite{rcText}. See Appendix~\ref{sec:fure} for further considerations of related work, with emphasis on the results from research herein reported.

\section{Data description}
    \subsection{Email lists and messages}
Email list messages were obtained from
the GMANE email archive~\cite{GMANE}, which consists of more than 20,000 email lists and more than 130,000,000 messages~\cite{GMANEwikipedia}. These lists cover a variety of topics, mostly technology-related. The archive can be described as a corpus with metadata of its messages, including sent time, place, sender name, sender email address.
GMANE usage in scientific research is reported in studies of isolated lists and of lexical innovations~\cite{GMANE2,bird}. The scripts for gathering and processing GMANE email messages are given in Appendix~\ref{scripts}.

    \subsection{Chosen dataset}
Four email lists were selected for their diversity, easing initial observance of natural and general properties.
\begin{itemize}
    \item Linux Audio Users list\footnote{gmane.linux.audio.users is list ID in GMANE.}. Dominated by participants with hybrid artistic and technological interests. Participants are from different countries, and English is the language used the most. Abbreviated as LAU from now on.
    \item Linux Audio Developers list\footnote{gmane.linux.audio.devel is list ID in GMANE.}. Participants are from different countries, and English is the language used the most. A more technical and less active version of LAU. Abbreviated LAD from now on.
    \item Development list for the standard C++ library\footnote{gmane.comp.gcc.libstdc++.devel is list ID in GMANE.}. Dominated by specialized computer programmers. Participants are from different countries, and English is the language used the most. Abbreviated as CPP from now on.
    \item List of the MetaReciclagem project\footnote{gmane.politics.organizations.metareciclagem is list ID in GMANE.}. Dominated by Brazilian activists and digital culture interests. Participants are mostly Brazilians, and Portuguese is the most used language, although Spanish and English are also incident. Abbreviated MET from now on.
\end{itemize}

 The first 20,000 messages of each list were considered, with total timespan, authors, threads and missing messages exposed in Table~\ref{geralListas}.

\begin{table}
  \centering
  \caption{Columns $date_1$ and $date_M$ have dates of first and last messages from the 20,000 messages considered in each email list.
$N$ is the number of participants (number of different email addresses).
$\Gamma$ is the number of threads (count of messages without antecedent).
$\overline{M}$ is the number of messages missing in the 20,000 collection, $100\frac{23}{20000}=0.115$ percent in the worst case.
MET notably has the fewer participants and the larger number of threads.
This relation holds for each of the lists considered: as the number of participants increases,
 the number of threads decreases.}
  \label{geralListas}
  \begin{tabular}{|l|c|c|c|c|c|}\hline
list & $date_1$ & $date_{M}$    & $N$  & $\Gamma$ & $\overline{M}$ \\\hline
LAU  & Jun/29/2003 & Jul/23/2005 & 1183 & 3373 & 5 \\
LAD  & Jun/30/2003 & Oct/07/2009 & 1268 & 3113 & 4 \\
MET  & Ago/01/2005 & Mar/07/2008 & 492  & 4607 & 23 \\
CPP  & Mar/13/2002 & Aug/25/2009 & 1052 & 4506 & 7 \\ \hline
  \end{tabular}
\end{table}

\section{Characterization methods}
After immersion with the data and appropriated literature, the following methods for network characterization were chosen: 1) statistics of activity along time; 2) division of network in hubs, intermediary and peripheral vertex; 3) prominence of topological measures; 4) evolutive visualizations and quantitative observations; 5) typological elaborations of networks and participants.
Each of these methods are described bellow with supporting structures.

    \subsection{Temporal activity statistics}
Number of messages along time with respect to seconds, minutes, hours, days of the week, days of the month, and months of the year. These are exhibited as tables in Appendix~\ref{tabTime}. Results are outlined in Section~\ref{constDisc}.

    \subsection{Interaction network}\label{intNet}
Regarding literature~\cite{bird,newmanCommunityDirected,newmanCommunity2013}, interaction networks can be modeled both weighted or unweighted, both directed or undirected. Networks in this article are directed and weighted, considered as more informative among possibilities (directed unweighted, undirected weighted, and undirected unweighted).
The networks were considered as follows: a direct response from participant B to a message from participant A yields an edge from A to B, as information went from A to B. The reasoning is: if B wrote a response to a message from A, he read what A wrote and formulated a response, so B assimilated information from A, thus $A \rightarrow B$. Inverting edge direction yields the status network, as B read the message and considered what A wrote worth responding, giving status to A, thus $B\rightarrow A$. This article uses the information network as described above and depicted in Figure~\ref{formationNetwork}. Edges in both directions are allowed. Each time an interaction occurs, one is added to edge weight. Self-loops were regarded as non-informative and discarded. This networks are described as exhibiting free-scale and small world properties, as expected for a social network~\cite{bird}.

\begin{figure}[hb]
    \centering
    \includegraphics[width=0.5\textwidth]{figs/criaRede_}
    \caption{Formation of interaction network from email messages. Each vertex represents a participant. A reply message from participant B to a message from participant A is regarded as evidence that B received information from A. Multiple messages add ``weight'' to directed edge. Further details are in Section~\ref{intNet}.}
    \label{formationNetwork}
\end{figure}

Previous messages on the thread create directed edges from their author to the observed message's author. Edges can be created from all antecedent messages on the message-response thread. In this work, only the immediate predecessor is linked to new message's author, both for simplicity and for the valid objection that in adding two edges, $x\rightarrow y$ and $y\rightarrow z$, there is also a connection between $x\rightarrow z$. Potential interpretations for this weaker connection are usually common sense, such as: double length, half weight or with one more ``obstacles''. This suggests the adoption of centrality measures that account for the connectivity with all nodes, such as betweenness centrality and accessibility~\cite{luMeasures,access}.

        \subsubsection{Sectioning network in periphery, intermediary and hubs classes}\label{sectioning}

\begin{figure}[hb]
    \centering
    \includegraphics[width=0.5\textwidth]{figs/fser}
    \caption{Degree distribution on scale-free and Erd\"os-R\'enyi ideal networks. The later has more
        intermediary, as the former has more peripheral and hub vertexes. Sections are
        given by the two intersections $k_1$ and $k_2$ of the connectivity distributions. Characteristic degrees
    are in compact intervals of degree: $[0,k_1]$, $(k_1,k_2]$, $(k_2,k_{max}]$ for the three sections considered (periphery, intermediary and hubs).}
    \label{fig:setores}
\end{figure}

Because of social networks tendency to have a scale-free distribution of connectivity,
one can compare it to an Erd\"os-R\'enyi random graph and consider peripheral, intermediary and hub
sectors~\cite{3setores}, as depicted in Figure~\ref{fig:setores}.
The degree distribution $\widetilde{P}(k)$ of an ideal
scale-free network $\mathcal{N}_f$ with $N$ vertexes and $z$ edges, has less
average degree nodes when compared with the distribution $P(k)$ of an Erd\"os-R\'enyi
random graph with the same number of vertexes and edges:

@@

If $\mathcal{N}_f$ is directed and has no self-loops, the probability
of the presence of an edge between two arbitrary vertexes is $p_e=\frac{z}{N(N-1)}$ (see Appendix~\ref{ap:ded}).
A vertex in the ideal Erd\"os-R\'enyi digraph with the same number of vertexes and edges, and thus the same probability $p_e$ for the presence of an edge, will have degree $k$ with probability:

@@

The lower degree fat tail is constituted by the border
vertexes and is called the peripheral sector. 
The higher degree fat tail is the hub sector.

The arguments behind this classification are: 1) vertexes so connected that they are virtually inexistent in networks whose edges are incident regardless of vertex properties, are correctly associated to the hubs sector. Vertexes with very few connections, which are way more abundant than expected in networks whose edges are incident regardless of vertex properties, are correctly assigned to periphery. Degree values near average, predicted as the most abundant if connections are created without criteria, and less frequent in free-scale phenomena, are classified to intermediary vertexes.

To ensure statistical validity, bins can be chosen to contain at least $\eta$ vertexes. Thus, each bin, starting at degree $k_i$, spans $\Delta_i=[k_{i},k_{j}]$ degree values, where $j$ is the smallest integer in which degrees $k_i$-$k_{j}$ contain at least $\eta$ vertexes. This changes equation~\ref{criterio} to:

@@

If instead strength $s$ is used for comparison, $P$ remains the same, but $P(\kappa_i)$ with $\kappa_i=\frac{s_i}{\overline{w}}$ should be used for comparison, with $\overline{w}=\frac{z}{\sum_is_i}$ the average weight of an edge and $s_i$ the strength of vertex $i$. For in and out degrees and strengths, comparisons should be made with $\kappa_i=2k_i^{in}$, $\kappa_i=2k_i^{out}$, $\kappa_i=2\frac{s_i^{in}}{\overline{w}}$ and $\kappa_i=2\frac{s_i^{out}}{\overline{w}}$. Results of these criteria for network segmentation are discussed in Section~\ref{subsec:pih}.

Compound criteria is also used for classification of vertex, considering all measures: total, in and out degree and strength. After a careful inspection of possible combinations, these were reduced to six:

\begin{itemize}
    \item Exclusivist criteria: vertexes are only classified if the class is the same according to all metrics. In this case, vertexes classified does not usually reach 100\%, which is depicted by a black line in Appendix~\ref{figures}.
    \item Inclusivist criteria: vertex has class given by any of the metrics. Therefore, a vertex can belong to more than one class, and class attribution may add more than total number of vertexes, which is indicated by a black line in Appendix~\ref{figures}.
    \item Exclusivist cascade: a hub is only classified as hub if it is a hub with respect to all metrics. Intermediary are the vertexes classified as intermediary or hub with respect to all metrics. Remaining vertexes are regarded as peripheral.
    \item Inclusivist cascade: hubs are vertexes classified as hubs by any of the metrics. The remaining vertexes are classified as intermediary if they belong to this category by any of the metrics. Vertexes only classified as peripheral are the only considered as such.
    \item Exclusivist externals: vertexes are only hubs if they are classified as such according to all the metrics. The remaining vertexes are classified as peripheral if they fall into hub or peripheral categories by any metric. The rest of the vertexes are classified as intermediary.
    \item Inclusivist externals: hubs are vertexes classified as hubs according to any metric. The remaining vertexes will be peripheral if classified as such according to any metric. The rest of the vertexes are regarded regarded as intermediary.
\end{itemize}

These compound criteria, and reduction of possibilities to them, can be formalized in strict mathematical terms. This was considered out of the scope of the present article. Important here is to notice that the compound criteria can be used to examine network sections in the case of a low number of messages, as will be shown later.

Results from applying this classification method is reported in Section~\ref{subsec:pih}.

        \subsubsection{Topological measures}\label{measures}
The topology of the networks was characterized by a small selection of the most standard
measurements of each vertex:

\begin{itemize}
    \item Degree $d_i$: number edges linked to                                 vertex $i$.
    \item In-degree $d_i^{in}$: number of edges ending at                      vertex $i$.
    \item Out-degree $d_i^{out}$: number of edges departing from               vertex $i$.
    \item Strength $s$: sum of weights of all edges linked to                  vertex $i$.
    \item In-strength $s_i^{in}$: sum of weights of all edges ending at        vertex $i$.
    \item Out-strength $s_i^{out}$: sum of weights of all edges departing from vertex $i$.
    \item Clustering coefficient $cc_i$: fraction of pairs of neighbors of vertex $i$ that are linked. This is the standard clustering coefficient for undirected graphs. Usage of other clustering coefficients (e.g. for directed graphs) was considering out of scope, as such clustering coefficients are usually employed in special contexts. 
    \item Betweenness centrality $bt_i$: fraction of geodesics that contain the vertex $i$. Betweenness centrality index considered directions and weight, as specified in~\cite{faster}.
\end{itemize}

In order to capture asymmetries in the activity of participants, the following metrics were introduced (see Section~\ref{prevalence}):

>>>>

\author{Renato Fabbri}%
 \homepage{http://ifsc.usp.br/~fabbri/}
 \email{fabbri@usp.br}
  \affiliation{ 
S\~ao Carlos Institute of Physics, University of S\~ao Paulo (IFSC/USP)%\\This line break forced with \textbackslash\textbackslash
}

\author{Vilson V. da Silva Jr.}
  \homepage{http://automata.cc/}
  \email{vilson@void.cc}
  \altaffiliation[Also at ]{IFSC-USP}%Lines break automatically or can be forced with \\

@@

\author{Deborah C. Antunes}
  \homepage{http://lattes.cnpq.br/1065956470701739}
  \email{deborahantunes@gmail.com}
  \altaffiliation{
Curso de Psicologia, Universidade Federal do Cer\'a (UFC)
}%Lines break automatically or can be forced with \\

\author{Marilia M. Pisani}
  \homepage{http://lattes.cnpq.br/6738980149860322}
  \email{marilia.m.pisani@gmail.com}
 \altaffiliation{
Centro de Ciências Naturais e Humanas, Universidade Federal do ABC (CCNH/UFABC)
}%Lines break automatically or can be forced with \\

%\author{Luciano da Fontoura Costa}
%  \homepage{http://cyvision.ifsc.usp.br/~luciano/}
%  \email{ldfcosta@gmail.com}
%  \altaffiliation[Also at ]{IFSC-USP}%Lines break automatically or can be forced with \\

\author{Osvaldo N. Oliveira Jr.}
  \homepage{www.polimeros.ifsc.usp.br/professors/professor.php?id=4}
  \email{chu@ifsc.usp.br}
 \altaffiliation[Also at ]{IFSC-USP}%Lines break automatically or can be forced with \\

@@

\begin{abstract}
 This article reports interaction networks stability by means of three quantitative criteria: activity distribution in time and among participants; a sound classification of vertices in peripheral, intermediary and hub sectors; the combination of basic measures into principal components with greater variance. We analyzed the temporal activity and topology evolution of networks in four email lists by considering  window sizes from 50 to 10,000 messages, which were made to slide to generate snapshots of the network along a timeline. Activity in terms of seconds, minutes, days and months was remarkably similar for all the networks. Participant activity follows concentrations expected in scale-free networks.  We compare these networks to Erd\"os-R\'enyi networks in order to assign members to three distinct sectors, namely hubs, intermediary and periphery. The fractions of members in these sectors were essentially the same for all networks and, most importantly, stable over time. If strength is used as the criterion for classification, for instance, ca. 5\% of the vertices are hubs, 15-20\% are intermediary and the remainder compose periphery. The metrics  most representative for  data dispersion were found to be centrality-related  (degree, strength and betweenness), followed by symmetry-related,  and then clustering coefficient. Degree, strength and clustering cope with different symmetry measures to bear secondary components. Again, the importance of these metrics to network topology dispersion was practically the same for all email lists examined, and stable over time. The research also resulted on a sketch of a physics-based typology of agents, with proper social and psychological speculations. Because the network properties reported did not depend on the email list and were stable over time, and because observed structure is in accordance with expectations driven from literature for human interaction networks, we believe that the properties observed  holds for general human interaction networks, and the approach can be applied to other types of networks. Current unfoldings include governance and accountability proposals and implementations, anthropological physics experimentations and the report of quantitative differences of textual production as connectivity of agents changes.
\end{abstract}

\pacs{89.75.Fb,05.65.+b,89.65.-s}% PACS, the Physics and Astronomy
\keywords{complex networks, social network analysis, pattern recognition, statistics, anthropological physics}%Use showkeys class option if keyword
\maketitle

%Obs. Em geral não se incluem citações como esta abaixo nos artigos, e podemos %ter problema de aceitação se um referee não entender o porquê de colocá-la.

\begin{quotation}
`The reason for the persistent plausibility of the typological approach, however, is not a static biological one, but just the opposite: dynamic and social. The fact that human society has been up to now divided into classes affects more than the external relations of men. The marks of social repression are left within the individual soul.'

% `The conception of personality structure is the
%best safeguard against the
% inclination to attribute persistent trends in the
% individual to something
% "innate" or "basic" or "racial" within him. The
% Nazi allegation that natural, biological traits decide the total being of a % person
% would not have been such
% a successful political device
% had it not been possible to point to numerous
% instances of relative fixity in human behavior and to
% challenge those who
% thought to explain them on any basis other than a biological one.'
\emph{- Adorno et al, 1969, p. 747}
\end{quotation}

@@

\subsection{Related work}

This work approaches network stability through temporal evolution observation.
This evolution is represented by a timeline of activity snapshots with a constant number of contiguous messages. These snapshots yield a succession of networks. This approach, although coherent and intuitive, seems not to be explored to date~\cite{netWindowEvo}.

Works on network evolution often consider solely network growth, in which there is a monotonic increase in the number of events considered~\cite{barabasiEvo}. Exceptions are reported in this section, specially those more closely related to the present article.

The evolution of human interaction networks was addressed with a community focus, where the direction of edges was not taken into account~\cite{barabasiEvo}. In another article, two topologically different networks emerged from human interaction networks, depending on the frequency of interactions, which could either be a generalized power law or an exponential connectivity distribution~\cite{barabasiTopologicalEv}. In email list networks, scale-free properties were verified~\cite{bird}, and different linguistic traces were related to weak and strong ties~\cite{GMANE2}. These results are corroborated by phenomena reported in this paper. See Appendix~\ref{sec:fure} for further considerations of related work, with emphasis on the results described below and deriving articles dedicated to texts produced by each of the primitive connective sectors~\cite{rcText}, a visualization method for network evolution~\cite{versinus} and an ongoing elaboration of the primitive typology.

\section{Data description}
    \subsection{Email lists and messages}
Email list messages were obtained from
the GMANE email archive~\cite{GMANE}, which consists of more than 20,000 email lists and more than 130,000,000 messages~\cite{GMANEwikipedia}. These lists cover a variety of topics, mostly technology-related. The archive can be described as a corpus with metadata of its messages, including sent time, place, sender name, and sender email address.
The GMANE usage in scientific research is reported in studies of isolated lists and of lexical innovations~\cite{GMANE2,bird}. The computer scripts for gathering and processing GMANE email messages are given in Appendix~\ref{scripts}.

Four email lists were selected for their diversity, making it easier to infer general properties.
\begin{itemize}
    \item Linux Audio Users list\footnote{gmane.linux.audio.users is list ID in GMANE.}. Dominated by participants with hybrid artistic and technological interests. Participants are from different countries, and English is the language used the most. Abbreviated as LAU from now on.
    \item Linux Audio Developers list\footnote{gmane.linux.audio.devel is list ID in GMANE.}. Participants are from different countries, and English is the language used the most. A more technical and less active version of LAU. Abbreviated LAD from now on.
    \item Development list for the standard C++ library\footnote{gmane.comp.gcc.libstdc++.devel is list ID in GMANE.}. Dominated by specialized computer programmers. Participants are from different countries, and English is the language used the most. Abbreviated as CPP from now on.
    \item List of the MetaReciclagem project\footnote{gmane.politics.organizations.metareciclagem is list ID in GMANE.}. Dominated by Brazilian activists and digital culture interests. Participants are mostly Brazilians, and Portuguese is the most used language, although Spanish and English are also incident. Abbreviated MET from now on.
\end{itemize}

 The first 20,000 messages of each list were considered, with total timespan, authors, threads and missing messages, as indicated in Table~\ref{geralListas}.

\begin{table}
  \centering
  \caption{Columns $date_1$ and $date_M$ have dates of first and last messages from the 20,000 messages considered in each email list.
$N$ is the number of participants (number of different email addresses).
$\Gamma$ is the number of threads (count of messages without antecedent).
$\overline{M}$ is the number of messages missing in the 20,000 collection, $100\frac{23}{20000}=0.115$ percent in the worst case.
MET notably has the fewer participants and the larger number of threads.
This relation holds for each of the lists considered here and a derived article (see Appendix~\ref{sec:fure}): as the number of participants increases, the number of threads decreases.}
  \label{geralListas}
  \begin{tabular}{|l|c|c|c|c|c|}\hline
list & $date_1$ & $date_{M}$    & $N$  & $\Gamma$ & $\overline{M}$ \\\hline
LAU  & Jun/29/2003 & Jul/23/2005 & 1183 & 3373 & 5 \\
LAD  & Jun/30/2003 & Oct/07/2009 & 1268 & 3113 & 4 \\
MET  & Ago/01/2005 & Mar/07/2008 & 492  & 4607 & 23 \\
CPP  & Mar/13/2002 & Aug/25/2009 & 1052 & 4506 & 7 \\ \hline
  \end{tabular}
\end{table}

\section{Characterization methods}
The email lists and the networks generated from them were characterized by: 1) statistics of activity along time, with a detailed analysis for time durations from seconds to years; 2)  sectioning of the networks in hubs, intermediary and peripheral vertices; 3) prominence of topological measures, including their time dependence; 4) iterative visualization and data mining of networks; 5) typological elaborations of networks and participants.  Each of these procedures are described below with supporting structures.

Distribution of activity among participants are in Table~\ref{autores} and indirectly through almost all results of this article, as degree and strength are highly correlated to activity.

  \subsection{Time activity statistics}                          
Messages were counted along time with respect to seconds, minutes, hours, days of the week, days of the month, and months of the year. These are exhibited as Tables~\ref{dia}-\ref{ano} in Appendix~\ref{tabTime}. Results are outlined in Section~\ref{constDisc}.

    \subsection{Interaction network}\label{intNet}                     
Regarding literature~\cite{bird,newmanCommunityDirected,newmanCommunity2013}, interaction networks can be modeled both weighted or unweighted, both directed or undirected. Networks in this article are directed and weighted, the more informative of trivial possibilities, i.e. we did not observe directed unweighted, undirected weighted, and undirected unweighted representations of the interaction network. The networks were obtained as follows: a direct response from participant B to a message from participant A yields an edge from A to B, as information went from A to B. The reasoning is: if B wrote a response to a message from A, he/she read what A wrote and formulated a response, so B assimilated information from A, thus $A \rightarrow B$. Inverting edge direction yields the status network: B read the message and considered what A wrote worth responding, giving status to A, thus $B\rightarrow A$. This article uses the information network as described above and depicted in Figure~\ref{formationNetwork}. Edges in both directions are allowed. Each time an interaction occurs, one is added to the edge weight. Self-loops were regarded as non-informative and discarded. These networks are reported in literature as exhibiting scale-free and small world properties, as expected for  (some) social networks~\cite{bird,newmanBook}.

\begin{figure}[hb]
    \centering
    \includegraphics[width=0.5\textwidth]{figs/criaRede_}
    \caption{Formation of interaction network from email messages. Each vertex represents a participant. A reply message from participant B to a message from participant A is regarded as evidence that B received information from A. Multiple messages add ``weight'' to a directed edge. Further details are given in Section~\ref{intNet}.}
    \label{formationNetwork}
\end{figure}

Edges can be created from all antecedent messages on the message-response thread. We only linked the immediate antecedent to the new message's author, both for simplicity and for the valid objection that in adding two edges, $x\rightarrow y$ and $y\rightarrow z$, there is also a weaker connection between $x$ and $z$. Potential interpretations for this weaker connection are: double length, half weight or with one more ``obstacles''. This suggests pertinence of centrality measures that account for the connectivity with all nodes, such as betweenness centrality and accessibility~\cite{luMeasures,access}.

        \subsubsection{Sectioning networks in periphery, intermediary and hubs sectors}\label{sectioning}

Social networks tend to have a scale-free distribution of connectivity, and the primitive sectors (periphery, intermediary and hubs) can be derived from comparison with an Erd\"os-R\'enyi  network with the same number of edges and vertices~\cite{3setores}, as depicted in Figure~\ref{fig:setores}.
The degree distribution $\widetilde{P}(k)$ of an ideal
scale-free network $\mathcal{N}_f$ with $N$ vertices and $z$ edges has less
average degree nodes than the distribution $P(k)$ of an Erd\"os-R\'enyi
network with the same number of vertices and edges. Indeed, we define in this work the intermediary sector of a network to be the set of all the nodes whose degree is less abundant in the real network than on the Erd\"os-R\'enyi model:

@@

If $\mathcal{N}_f$ is directed and has no self-loops, the probability
of an edge between two arbitrary vertices is $p_e=\frac{z}{N(N-1)}$ (see Appendix~\ref{ap:ded}).
A vertex in the ideal Erd\"os-R\'enyi digraph with the same number of vertices and edges, and thus the same probability $p_e$ for the presence of an edge, will have degree $k$ with probability:

@@

The lower degree fat tail represents the border vertices, i.e. the peripheral sector or periphery where $\widetilde{P}(k)>P(k)$ and $k$ is lower and any intermediary sector value of $k$. The higher degree fat tail is the hub sector, i.e. $\widetilde{P}(k)>P(k)$ and $k$ is higher than any intermediary sector value of $k$. The reasoning for this classification is: 1) vertices so connected that they are virtually inexistent in networks connected at pure chance (e.g. without preferential attachment) are correctly associated to the hubs sector. Vertices with very few connections, which are way more abundant than expected by pure chance, are assigned to the periphery. Vertices with degree values predicted as the most abundant if connections are created by pure chance, near the average, and less frequent in scale-free phenomena, are classified as intermediary. 
\begin{figure}[!h]
    \centering
    \includegraphics[width=0.5\textwidth]{figs/fser}
    \caption{Degree distribution on scale-free and Erd\"os-R\'enyi ideal networks. The latter has more
        intermediary vertices, while the former has more peripheral and hub vertices. Sector borders are
        given by the two intersections $k_1$ and $k_2$ of the connectivity distributions. Characteristic degrees
    are in compact intervals of degree: $[0,k_1]$, $(k_1,k_2]$, $(k_2,k_{max}]$ for the three sectors considered (periphery, intermediary and hubs).}
    \label{fig:setores}
\end{figure}

To ensure statistical validity, bins can be chosen to contain at least $\eta$ vertices. Thus, each bin, starting at degree $k_i$, spans $\Delta_i=[k_{i},k_{j}]$ degree values, where $j$ is the smallest integer with which there are at least $\eta$ vertices with degree larger than or equal $k_i$, and less than or equal $k_{j}$. This changes equation~\ref{criterio} to:

@@

If strength $s$ is used for comparison, $P$ remains the same, but $P(\kappa_i)$ with $\kappa_i=\frac{s_i}{\overline{w}}$ should be used for comparison, with $\overline{w}=2\frac{z}{\sum_is_i}$ the average weight of an edge and $s_i$ the strength of vertex $i$. For in and out degrees and strengths, comparison should be made with $\kappa_i=2k_i^{in}$, $\kappa_i=2k_i^{out}$, $\kappa_i=2\frac{s_i^{in}}{\overline{w}}$ and $\kappa_i=2\frac{s_i^{out}}{\overline{w}}$. Results of these criteria for network segmentation are discussed in Section~\ref{subsec:pih} and exhibited in Figures~\ref{fig:cpp10000} to~\ref{fig:lad50_} of Appendix~\ref{figures}.

Since different metrics can be used in the segmentation to identify the three types of vertices, various criteria can be defined, e.g. with a very stringent criterion according to which a vertex will only be classified as hub if it is so for all the metrics. After a careful inspection of possible combinations, these were reduced to six:

\begin{itemize}
\item Exclusivist criterion:  vertices are only classified if the class is the same according to all metrics. In this case, vertices classified (usually) does not reach 100\%, which is indicated by a black line in the figures of Appendix~\ref{figures}.
        \item Inclusivist criterion: a vertex has the class given by any of the metrics. Therefore, a vertex can belong to more than one class, and the percentages of members may add to more than 100\%, which is indicated by a black line in the figures of Appendix~\ref{figures}.
    \item Exclusivist cascade: vertices are only classified as hubs if they are hubs according to all metrics. Intermediary are the vertices classified either as intermediary or hubs with respect to all metrics. The remaining vertices are regarded as peripheral.
    \item Inclusivist cascade: vertices are hubs if they are so classified according to any of the metrics. The remaining vertices are classified as intermediary, if they belong to this category for any of the metrics. Peripheral vertices will then be those which were not classified as hub or intermediary with any of the metrics. 
    \item Exclusivist externals: vertices are only hubs if they are classified as such according to all the metrics. The remaining vertices are classified as peripheral if they fall into the periphery or hub classes by any metric. The rest of the nodes are classified as intermediary.
    \item Inclusivist externals: hubs are vertices classified as hubs according to any metric. The remaining vertices will be peripheral if they are classified as such according to any metric. The rest of the vertices will be intermediary vertices.
\end{itemize}

These compound criteria, and the simplification of possibilities listed above (exclusivist/inclusivist, criterion/cascade/externals) , can be formalized in strict mathematical terms, but this was considered out of the scope of the present article. Important here is to notice that the compound criteria can be used to examine network sections in the case of a low number of messages, such as in the last figures of Appendix~\ref{figures}.

Results from applying this classification method, i.e. the achievement of well defined sectors by comparison of the real network with the Erd\"os R\'enyi model, are reported in Section~\ref{subsec:pih} and Appendix~\ref{figures}.

\subsubsection{Topological measurements}\label{measures}

The topology of the networks was characterized with a small selection of the most standard measurements for each vertex, as follows:

\begin{itemize}
    \item Degree $k_i$: number edges linked to vertex $i$.
    \item In-degree $k_i^{in}$: number of edges ending at vertex $i$.
    \item Out-degree $k_i^{out}$: number of edges departing from vertex $i$.
    \item Strength $s$: sum of weights of all edges linked to vertex $i$.
    \item In-strength $s_i^{in}$: sum of weights of all edges ending at vertex $i$.
    \item Out-strength $s_i^{out}$: sum of weights of all edges departing from vertex $i$.
    \item Clustering coefficient $cc_i$: fraction of pairs of neighbors of $i$ that are linked.  The standard clustering coefficient for undirected graphs was used.
  \item Betweenness centrality $bt_i$: fraction of geodesics that contain the vertex $i$. Betweenness centrality index considered directions and weight, as specified in~\cite{faster}.
\end{itemize}

In order to capture symmetries in the activity of participants, the following metrics were introduced for a vertex $i$ (see Section~\ref{prevalence}):

\begin{itemize}
    \item Asymmetry: $asy_i=\frac{d_i^{in}-d_i^{out}}{d_i}$.
    \item Mean of asymmetry of edges: $\mu_i^{asy}=\frac{\sum_{j\in J_i} e_{ji}-e_{ij}}{|J_i|}$. Where $e_{xy}$ is 1 if there is and edge from $x$ to $y$, $0$ otherwise. $|J_i|$ is the number of neighbors of vertex $i$.
    \item Standard deviation of asymmetry of edges: $\sigma_i^{asy}=\sqrt{\frac{\sum_{j\in J_i}[\mu_{asy} -(e_{ji}-e_{ij}) ]^2  }{|J_i|}  }$
    \item Disequilibrium: $dis_i=\frac{s_i^{in}-s_i^{out}}{s_i}$.
    \item Mean of disequilibrium of edges: $\mu_i^{dis}=\frac{\sum_{j \in J_i}\frac{w_{ji}-w_{ij}}{s_i}}{|J_i|}$, where $w_{xy}$ is the weight of edge $x\rightarrow y$ and zero if there is no such edge.
    \item Standard deviation of disequilibrium of edges: $\sigma_i^{dis}=\sqrt{\frac{\sum_{j\in J_i}[\mu_{dis}-\frac{(w_{ji}-w_{ij})}{s_i}]^2}{|J_i|}}$
\end{itemize}

   \subsection{Evolution of the networks}
The evolution of the networks was observed within a fixed number of messages (or window size $ws$) that shifts in the message timeline.
The $ws$ used were 50, 100, 200, 400, 500, 800, 1000, 2000, 2500, 5000 and 10000. Within a same $ws$, the number of vertices and edges vary in time, as do other network characteristics.
